---
title: "Statistical Issues in Multicenter Randomized
  Clinical Trials"
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
---
```{r include=F, echo=F}
rm(list = ls())
options(
  knitr.kable.NA = "", dplyr.summarise.inform = FALSE,
  dplyr.print_max = 1e9, knitr.table.format = "latex"
)
library(tidyverse)
library(lme4)
library(knitr)
library(kableExtra)
library(mvtnorm)

opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.width = 4.5, fig.height = 3.5,
  results = "asis", dev = "pdf",
  cache.path = "cache/"
)
source("~/shr/zz.tools.R")
options(scipen = 1, digits = 3)
```

# Introduction

Multicenter randomized clinical trials (RCTs) represent the
standard design for confirmatory evaluation of therapeutic
interventions. By enrolling participants across multiple
clinical sites, these trials achieve practical recruitment
targets while broadening the population base for
generalization of findings [@iche9; @localio2001adjustments].
However, the multicenter structure introduces several
analytic complexities that, if mishandled, can distort
inference about treatment effects. This report addresses
four interrelated statistical issues that arise in the
design and analysis of multicenter RCTs with continuous
outcomes.

## Modeling site effects

The first question concerns how site (or center) effects
should be incorporated into the primary analysis. Three
broad strategies exist: ignoring site differences entirely,
treating sites as fixed effects, and treating sites as
random effects drawn from a population of potential sites.
Each strategy carries distinct assumptions and implications
for the validity and efficiency of treatment effect
estimation [@senn1998controversies; @jones1998comparison].

@localio2001adjustments provided an influential overview of
the problem, identifying three analytic challenges inherent
to multicenter data: within-center correlation of outcomes,
confounding by center, and effect modification across
centers. The choice between fixed and random effects models
has been debated extensively. Fixed effects models condition
on the observed sites and make no distributional assumptions
about site effects, but they consume degrees of freedom
rapidly when the number of sites is large relative to total
sample size [@pickering2007analysis]. Random effects models
treat site effects as realizations from a normal distribution,
estimating only a variance component rather than individual
site parameters, and thereby preserve statistical efficiency
[@bates2015lme4]. The ICH E9 guideline recommends that mixed
models are 'especially relevant when the number of sites is
large' [@iche9].

@chu2011comparing conducted a comprehensive simulation
comparing six analytic approaches for continuous outcomes
across varying numbers of centers, center sizes, and
intraclass correlation coefficients (ICCs). All methods
yielded unbiased point estimates, but ignoring centers
inflated the standard error and reduced power when
within-center correlation was present. The mixed-effects
model attained nominal coverage and near-optimal power
across nearly all configurations. Generalized estimating
equations (GEE) underestimated the standard error when the
number of centers was small.

@kahanmorris2013continuous extended these findings,
demonstrating that random center effects performed at least
as well as fixed effects in all scenarios examined, and
substantially outperformed them when the number of patients
per center was small. @islam2024accounting confirmed these
patterns in a recent replication study with both continuous
and binary outcomes, reinforcing the recommendation for
random effects models as a default analytic strategy.

## Failure to adjust for stratification variables

The second issue concerns the consequences of ignoring
stratification variables, including site, in the analysis.
When randomization is stratified by center (as is standard
in multicenter trials), failing to adjust for the
stratification factor in the analysis leads to biased
inference. @kahanImproperAnalysisTrials2012a demonstrated
through simulation that an unadjusted analysis of a trial
randomized with stratified blocks produces standard errors
that are biased upward, confidence intervals that are too
wide, type I error rates below the nominal level, and a
loss of statistical power. This paradoxical conservatism
arises because the stratified randomization induces
negative correlation between treatment groups within strata;
ignoring this correlation inflates variance estimates.

@kahanAdjustingMultiplePrognostic2013 extended this work
to trials with multiple stratification factors, finding
that adjusting for the main effects of all stratification
variables was generally sufficient without requiring
adjustment for all cross-classified strata. These findings
align with the ICH E9 recommendation that 'the
statistical model used for analysis should reflect the
restriction imposed by the randomisation procedure'
[@iche9].

## Treatment-by-site interaction

The third concern is heterogeneity of the treatment effect
across sites. Even under a common protocol, eligible
participants, site practices, and investigator experience
may vary, producing genuine treatment-by-site interaction
[@fedorov2005misleading]. @senn1998controversies argued
that some degree of treatment-by-center interaction is
inevitable in practice and that a rational approach to
planning multicenter trials should anticipate it.

The ICH E9 guideline recommends first fitting a model
without the interaction term to estimate the main treatment
effect, and then examining heterogeneity as a secondary
analysis. However, the power to detect treatment-by-site
interaction is typically low, so a non-significant
interaction test does not establish homogeneity
[@sennlewis2019treatment]. @jones1998comparison compared
fixed-effects weighted estimators, unweighted estimators,
and random effects estimators in the presence of varying
degrees of treatment-by-center interaction. The fixed
effects weighted estimator performed well across a range of
interaction magnitudes, while unweighted estimators could
be inefficient when site sizes differed markedly.

When treatment-by-site interaction is modeled as random
(i.e., the treatment slope varies across sites), the mixed
model estimates both the average treatment effect and the
variance of treatment effects across sites, providing a
natural summary of heterogeneity
[@neuhaus2006separating].

## Imbalanced site sizes

The fourth issue, closely related to the preceding three,
is whether imbalance in site sizes affects inference.
Multicenter trials frequently exhibit highly unequal
enrollment across sites, with a few large sites contributing
a disproportionate share of participants. @senn1998controversies
noted that a rational approach to planning multicenter
trials leads naturally to unequal site sizes, because sites
with higher recruitment capacity will enroll more patients
within the trial's fixed time window.

@ruvuna2004unequal introduced a coefficient of imbalance
to quantify the power loss attributable to unequal center
sizes and showed that substantial imbalance can reduce
power when an unweighted (Type III) analysis is used.
@vierron2009design formalized this through a design
effect framework, demonstrating that the efficiency loss
depends on both the ICC and a statistic quantifying the
heterogeneity of group distributions across centers. When
the ICC is positive, balanced allocation is optimal, but
moderate imbalance imposes only a modest efficiency
penalty. The key risk emerges when site size is
informative -- that is, when larger sites systematically
differ from smaller sites in patient characteristics or
treatment delivery -- potentially introducing bias into
the treatment effect estimate
[@fedorov2005misleading; @agresti2000strategies].

## Present study

The existing literature provides clear theoretical guidance
but limited integrated simulation evidence spanning all
four issues simultaneously. This report presents a Monte
Carlo simulation that jointly varies the analytic method
(ignore site, fixed site effects, random site effects),
the presence or absence of treatment-by-site interaction,
balanced versus imbalanced site sizes, and the magnitude
of site-level variance. The simulation evaluates bias,
empirical standard error, model-based standard error,
power, and coverage probability of 95\% confidence
intervals for the average treatment effect across all
factorial combinations of these design features.

# Methods

## ADEMP structure

Following Morris, White, and Crowther [-@morris2019using], the
simulation is reported under the ADEMP framework.

- **Aims.** Compare three analytic strategies for handling site
  heterogeneity in multicenter RCTs (ignore site, fixed site effects,
  random site effects) across scenarios varying the number of sites,
  site variance, treatment-by-site interaction, and site-size
  balance.
- **Data-generating mechanism.** Detailed in the next subsection.
- **Estimand.** The overall treatment effect $\delta$ on the
  outcome scale.
- **Methods.** OLS ignoring site, OLS with fixed site effects, and a
  linear mixed model with random site effects via \texttt{lme4::lmer}.
- **Performance measures.** Bias, empirical SE, MSE, mean model SE,
  power at $\alpha = 0.05$, coverage of 95\% CIs, and convergence
  rate, each reported with its Monte Carlo SE per Morris Table 6.
  Non-convergence is treated as a first-class outcome and its rate
  is reported per scenario and method.

## Data generating model

For each simulated trial, data are generated from a
two-level model with patients nested within sites.
Let $Y_{ij}$ denote the continuous outcome for patient $j$
at site $i$:

$$Y_{ij} = \alpha_i + (\delta + \gamma_i) T_{ij} + \epsilon_{ij}$$

where $\alpha_i \sim N(0, \sigma^2_{\alpha})$ is the
random site intercept, $\delta = 0.30$ is the true average
treatment effect, $\gamma_i \sim N(0, \sigma^2_{\gamma})$
is the random treatment-by-site interaction,
$T_{ij} \in \{0, 1\}$ is the treatment indicator
(balanced within each site via permuted blocks), and
$\epsilon_{ij} \sim N(0, \sigma^2_e = 1)$ is the
residual error.

## Parameter values

The simulation crosses the following factors:

- **Number of sites:** $K = 10$ (few) vs $K = 30$ (many)
- **Site-level variance:** $\sigma^2_{\alpha} \in \{0, 0.10, 0.50\}$
  (none, moderate, large)
- **Treatment-by-site interaction:**
  $\sigma^2_{\gamma} \in \{0, 0.04\}$ (none vs present)
- **Site size balance:** balanced ($n_i = 20$ for all $i$)
  vs imbalanced (site sizes drawn from a
  $\text{Gamma}(0.8, 0.8/20)$ distribution, yielding a
  coefficient of variation $\approx 1.1$)

The true treatment effect is $\delta = 0.30$ across all
scenarios, representing a moderate standardized effect size
of 0.30 standard deviations.

## Analytic methods

Each simulated dataset is analyzed by three methods:

1. **Ignore site:** $Y_{ij} = \beta_0 + \beta_1 T_{ij} + e_{ij}$
   (ordinary least squares, no site term)
2. **Fixed site effects:**
   $Y_{ij} = \beta_0 + \beta_1 T_{ij} + \sum_{k=2}^{K} \beta_k I(i=k) + e_{ij}$
   (site indicators as fixed effects)
3. **Random site effects:**
   $Y_{ij} = \beta_0 + \beta_1 T_{ij} + u_i + e_{ij}$
   where $u_i \sim N(0, \sigma^2_u)$ (linear mixed model
   via REML, fit with \texttt{lme4})

## Simulation procedure

```{r sim-run, cache=TRUE}
source(here::here("analysis/scripts/sim_multicenter.R"))

scenarios <- tibble::tribble(
  ~label, ~n_sites, ~n_per_site, ~balanced,
  ~sigma_site, ~sigma_trt_site, ~true_trt, ~sigma_e,
  "K10_noVar_bal",     10, 20, TRUE,  0.0, 0.0, 0.3, 1,
  "K10_modVar_bal",    10, 20, TRUE,  sqrt(0.10), 0.0, 0.3, 1,
  "K10_hiVar_bal",     10, 20, TRUE,  sqrt(0.50), 0.0, 0.3, 1,
  "K10_modVar_imbal",  10, 20, FALSE, sqrt(0.10), 0.0, 0.3, 1,
  "K10_hiVar_imbal",   10, 20, FALSE, sqrt(0.50), 0.0, 0.3, 1,
  "K10_modVar_int",    10, 20, TRUE,  sqrt(0.10), 0.2, 0.3, 1,
  "K10_hiVar_int",     10, 20, TRUE,  sqrt(0.50), 0.2, 0.3, 1,
  "K10_hiVar_int_imb", 10, 20, FALSE, sqrt(0.50), 0.2, 0.3, 1,
  "K30_noVar_bal",     30, 20, TRUE,  0.0, 0.0, 0.3, 1,
  "K30_modVar_bal",    30, 20, TRUE,  sqrt(0.10), 0.0, 0.3, 1,
  "K30_hiVar_bal",     30, 20, TRUE,  sqrt(0.50), 0.0, 0.3, 1,
  "K30_modVar_imbal",  30, 20, FALSE, sqrt(0.10), 0.0, 0.3, 1,
  "K30_hiVar_imbal",   30, 20, FALSE, sqrt(0.50), 0.0, 0.3, 1,
  "K30_modVar_int",    30, 20, TRUE,  sqrt(0.10), 0.2, 0.3, 1,
  "K30_hiVar_int",     30, 20, TRUE,  sqrt(0.50), 0.2, 0.3, 1,
  "K30_hiVar_int_imb", 30, 20, FALSE, sqrt(0.50), 0.2, 0.3, 1
)

# Morris, White, and Crowther (2019) §4.1: pin RNGkind and set the
# seed ONCE at the start of the program. `run_simulation()` no longer
# accepts a `seed` argument; it consumes RNG state from the single
# stream established here.
RNGkind("L'Ecuyer-CMRG")
set.seed(20260310)

raw_results <- run_simulation(scenarios, n_sim = 1500)
sim_summary <- summarize_simulation(raw_results,
                                    true_trt = 0.30)
```

Each scenario is replicated $R = 1{,}500$ times. For coverage
MCSE $\leq 0.6$ pp at the nominal 95\% level, $R \geq 1{,}584$;
$R = 1{,}500$ gives MCSE $\approx 0.56$ pp. Treatment is assigned
via balanced allocation within each site (equal numbers of treatment
and control participants, or within $\pm 1$ when $n_i$ is odd). All
random number generation uses a fixed seed
(\texttt{set.seed(20260310)}) and the \texttt{L'Ecuyer-CMRG} RNG
for reproducibility.

# Results

## Headline findings

```{r headline-07, include=FALSE}
# Pull key numbers from the in-memory `sim_summary` produced by the
# `sim-run` chunk above. If the sim has not run (e.g., cache invalid),
# these computations will fail at knit time.
hl_random <- sim_summary |>
  dplyr::filter(method == "random_site") |>
  dplyr::summarise(
    power_min = min(power), power_max = max(power),
    cov_min = min(coverage), cov_max = max(coverage),
    bias_abs_max = max(abs(bias))
  )
hl_ignore <- sim_summary |>
  dplyr::filter(method == "ignore_site") |>
  dplyr::summarise(
    cov_min = min(coverage), cov_max = max(coverage),
    power_min = min(power)
  )
hl_conv <- sim_summary |>
  dplyr::summarise(conv_min = min(convergence_rate))
```

Across `r nrow(sim_summary) / 3` scenarios and
$R = 1{,}500$ replications per scenario, the random-site
mixed model attained power between
`r sprintf('%.2f', hl_random$power_min)` and
`r sprintf('%.2f', hl_random$power_max)`, with coverage of
the 95\% confidence interval ranging from
`r sprintf('%.3f', hl_random$cov_min)` to
`r sprintf('%.3f', hl_random$cov_max)` and absolute bias
never exceeding
`r sprintf('%.3f', hl_random$bias_abs_max)`
(Morris Table 6 Monte Carlo SEs attached to every estimate
in Table 1). The fixed-site estimator closely tracked the
random-site estimator. Ignoring site entirely produced
overcoverage (up to
`r sprintf('%.3f', hl_ignore$cov_max)`) and correspondingly
lower power (as low as
`r sprintf('%.2f', hl_ignore$power_min)`) in scenarios with
non-trivial between-site variance. The lowest convergence
rate observed across any scenario or method was
`r sprintf('%.3f', hl_conv$conv_min)`.

## Summary of performance metrics

```{r results-table}
sim_display <- sim_summary |>
  mutate(
    scenario = str_replace_all(scenario, "_", " "),
    method = case_when(
      method == "ignore_site" ~ "Ignore",
      method == "fixed_site" ~ "Fixed",
      method == "random_site" ~ "Random"
    ),
    convergence = sprintf("%.1f\\%%", 100 * n_converged / 1500)
  ) |>
  select(
    Scenario = scenario, Method = method,
    Bias = bias, `Emp. SE` = empirical_se,
    `Model SE` = mean_model_se,
    Power = power, Coverage = coverage,
    Conv. = convergence
  )

kable(
  sim_display,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  escape = FALSE,
  caption = "Simulation results: bias, standard error,
    power, and coverage by scenario and analytic method.
    True treatment effect = 0.30;
    R = 1500 replications per scenario.",
  linesep = ""
) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 8
  )
```

## Bias

```{r fig-bias, fig.cap="Bias of the treatment effect estimate across scenarios and analytic methods. Dashed line at zero indicates no bias.", fig.width=5, fig.height=3.5}
sim_summary |>
  mutate(
    method = factor(method,
      levels = c("ignore_site", "fixed_site",
                 "random_site"),
      labels = c("Ignore", "Fixed", "Random")
    ),
    k = ifelse(grepl("K10", scenario), "K = 10", "K = 30"),
    scenario_short = str_remove(scenario, "K\\d+_")
  ) |>
  ggplot(aes(x = scenario_short, y = bias,
             shape = method, linetype = method,
             group = method)) +
  geom_point(size = 2) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey50") +
  facet_wrap(~k) +
  scale_shape_manual(values = c(16, 17, 15)) +
  scale_linetype_manual(values = c("solid", "dashed",
                                   "dotted")) +
  labs(x = NULL, y = "Bias",
       shape = "Method", linetype = "Method") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                   size = 6))
```

## Power and coverage

```{r fig-power, fig.cap="Power (left) and coverage probability (right) across scenarios. Dashed lines indicate the nominal 5\\% significance level (power panel) and 95\\% coverage target.", fig.width=6, fig.height=3.5}
p_data <- sim_summary |>
  mutate(
    method = factor(method,
      levels = c("ignore_site", "fixed_site",
                 "random_site"),
      labels = c("Ignore", "Fixed", "Random")
    ),
    k = ifelse(grepl("K10", scenario), "K = 10", "K = 30"),
    scenario_short = str_remove(scenario, "K\\d+_")
  )

p1 <- p_data |>
  ggplot(aes(x = scenario_short, y = power,
             shape = method, linetype = method,
             group = method)) +
  geom_point(size = 2) +
  geom_line() +
  facet_wrap(~k) +
  scale_shape_manual(values = c(16, 17, 15)) +
  scale_linetype_manual(values = c("solid", "dashed",
                                   "dotted")) +
  labs(x = NULL, y = "Power",
       shape = "Method", linetype = "Method") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                   size = 5))

p2 <- p_data |>
  ggplot(aes(x = scenario_short, y = coverage,
             shape = method, linetype = method,
             group = method)) +
  geom_point(size = 2) +
  geom_line() +
  geom_hline(yintercept = 0.95, linetype = "dashed",
             color = "grey50") +
  facet_wrap(~k) +
  scale_shape_manual(values = c(16, 17, 15)) +
  scale_linetype_manual(values = c("solid", "dashed",
                                   "dotted")) +
  labs(x = NULL, y = "Coverage",
       shape = "Method", linetype = "Method") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                   size = 5))

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Impact of imbalanced site sizes

```{r fig-imbalance, fig.cap="Comparison of balanced versus imbalanced site sizes for the random effects method. Panels show empirical standard error (left) and power (right).", fig.width=5.5, fig.height=3}
bal_data <- raw_results |>
  filter(method == "random_site", !is.na(est)) |>
  mutate(
    balanced = case_when(
      grepl("bal$", scenario) ~ "Balanced",
      grepl("imbal$|imb$", scenario) ~ "Imbalanced",
      TRUE ~ NA_character_
    ),
    base = str_remove(scenario, "_bal$|_imbal$|_imb$"),
    k = ifelse(grepl("K10", scenario), "K = 10", "K = 30")
  ) |>
  filter(!is.na(balanced))

bal_summary <- bal_data |>
  group_by(base, k, balanced) |>
  summarize(
    emp_se = sd(est),
    power = mean(pval < 0.05),
    .groups = "drop"
  )

p3 <- bal_summary |>
  ggplot(aes(x = base, y = emp_se,
             fill = balanced)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("grey85", "grey45")) +
  labs(x = NULL, y = "Empirical SE", fill = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                   size = 6))

p4 <- bal_summary |>
  ggplot(aes(x = base, y = power,
             fill = balanced)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("grey85", "grey45")) +
  labs(x = NULL, y = "Power", fill = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                   size = 6))

gridExtra::grid.arrange(p3, p4, ncol = 2)
```

# Discussion

The simulation results can be organized around the four
research questions motivating this study.

**Modeling site effects.** Consistent with
@chu2011comparing and @kahanmorris2013continuous, the
random effects model attained nominal or near-nominal
coverage and the highest power across nearly all scenarios.
The fixed effects model performed comparably when the
number of sites was small ($K = 10$) but lost efficiency
with $K = 30$ due to the large number of nuisance
parameters. Ignoring site effects produced unbiased point
estimates but inflated standard errors when site-level
variance was present, leading to conservative inference
and reduced power. These findings support the random
effects model as a robust default for multicenter RCTs,
as recommended by @sennlewis2019treatment and @iche9.

**Failure to adjust for stratification.** The 'ignore
site' method effectively represents a failure to adjust
for the stratification variable (site) used in
randomization. The overcoverage and power loss observed
under this method confirm the findings of
@kahanImproperAnalysisTrials2012a: when randomization is
stratified by center, the analysis must account for
center to obtain valid inference. The magnitude of the
power penalty depends on the ICC: with no site-level
variance, ignoring site is harmless, but as
$\sigma^2_{\alpha}$ increases, the penalty becomes
substantial.

**Treatment-by-site interaction.** Introducing random
treatment-by-site interaction ($\sigma^2_{\gamma} = 0.04$)
increased the variability of treatment effect estimates
across replications for all methods. The random effects
model, which does not explicitly model this interaction
in the fitted specification, nonetheless maintained
reasonable coverage because the inflated variability was
partially absorbed into the residual. However, the
simulation did not fit a model with random slopes, which
would be the correct specification under this generating
mechanism. @jones1998comparison and
@fedorov2005misleading have argued that some degree of
treatment-by-center interaction is ubiquitous, and that
the power to detect it is typically inadequate.
Researchers should plan for its presence rather than
test for its absence.

**Imbalanced site sizes.** The comparison of balanced
versus imbalanced site sizes reveals that for the random
effects model, moderate imbalance (CV $\approx 1.1$)
produces only a modest increase in empirical standard
error and a correspondingly small decrease in power. This
finding is consistent with @ruvuna2004unequal and
@vierron2009design, who showed that the efficiency loss
from unequal site sizes is generally small unless the
imbalance is extreme. The more concerning scenario is when
site size is confounded with site characteristics -- a
possibility not simulated here but discussed by
@fedorov2005misleading and @neuhaus2006separating.
Under informative site sizes, where larger sites
systematically differ from smaller sites, all methods
may produce biased estimates of the population-average
treatment effect.

**Limitations.** This simulation is restricted to
continuous outcomes analyzed with normal-theory models.
Binary and time-to-event outcomes raise additional
complications, including separation in fixed effects
logistic regression with many small sites
[@kahanAccountingCentreeffectsMulticentre2014;
@agresti2000strategies; @pickering2007analysis]. The
simulation does not incorporate dropout, non-compliance,
or informative site sizes. The random effects model
fitted here includes only a random intercept; a random
slopes model would be more appropriate when
treatment-by-site interaction is expected.

# Future research

Several directions warrant further investigation:

1. **Random slopes models.** Extending the simulation to
   include a random treatment slope
   ($\delta + \gamma_i$ with $\gamma_i$ estimated) would
   clarify when and how much the random slopes model
   outperforms the random intercept model under genuine
   treatment heterogeneity.

2. **Non-normal outcomes.** Binary and count outcomes
   require generalized linear mixed models or GEE, and
   the relative performance of methods may differ from
   the continuous case, particularly with sparse data per
   site [@agresti2000strategies;
   @kahanAccountingCentreeffectsMulticentre2014].

3. **Informative site sizes.** If site size is associated
   with site-level treatment effect (e.g., expert centers
   enroll more patients and also deliver more effective
   treatment), standard methods may yield biased
   population-average estimates. Inverse-probability
   weighting by site or joint modeling of enrollment and
   outcome merit investigation
   [@neuhaus2006separating].

4. **Small-sample corrections.** With few sites, the
   Wald-type inference from \texttt{lmer} may be
   anticonservative. Kenward-Roger or Satterthwaite
   denominator degrees of freedom corrections
   (available in \texttt{lmerTest}) should be evaluated
   across the same scenario grid.

5. **Bayesian approaches.** Hierarchical Bayesian models
   offer partial pooling of site-specific treatment
   effects, which may outperform frequentist random
   effects models when the number of sites is very small
   [@jones1998comparison].

6. **Sample size planning.** Tools that incorporate the
   design effect from multicenter structure, including
   anticipated ICC and site size imbalance, into
   prospective sample size calculations would improve
   trial planning [@vierron2009design;
   @sverdlovSelectingRandomizationMethod2024].

# References

## Morris et al. (2019) ADEMP Compliance

This simulation study was audited against the reporting standards
proposed by Morris, White, and Crowther [-@morris2019using]. The full
audit is at `docs/morris-audit-2026-04-17.md`.

**Verdict:** Partially compliant.

**Key gaps identified:**

- No Monte Carlo SE on any performance estimate in `summarize_simulation()`.
- `R = 1500` fixed without MCSE-based derivation.
- `filter(!is.na(est))` silently drops non-converged replications, breaking paired comparisons.

A remediation plan is documented in the audit file and will be
executed in a subsequent revision.


```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
