---
title: "Statistical Issues in Multicenter Randomized
  Clinical Trials"
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
# The 5cm right margin (vs. 3cm left) is intentional, not an error:
# it is reserved as annotation space alongside the \linenumbers
# output below, for reviewer/self-review markup during drafting.
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
abstract: |
  Multicenter randomized controlled trials require a choice among
  ignoring, fixing, or randomizing site effects in the primary
  analysis, and this choice interacts with stratified-randomization
  adjustment, treatment-by-site interaction, and site-size balance.
  We report a factorial Monte Carlo simulation, structured under the
  ADEMP framework of Morris, White, and Crowther (2019), that jointly
  varies the number of sites, site-level variance, treatment-by-site
  interaction, and site-size balance, comparing OLS ignoring site,
  OLS with fixed site effects, and a linear mixed model with random
  site effects. In addition to bias, empirical and model-based
  standard error, power, and coverage under a true alternative
  treatment effect, we report empirical Type I error under a true
  null effect, and singular/boundary-fit rates for the random-effects
  model, both previously unreported in this simulation design. The
  random-effects model attains the best combination of power and
  coverage across most conditions; ignoring site produces conservative
  (not anti-conservative) inference under stratified randomization,
  consistent with prior work; and all three methods control Type I
  error near the nominal level in the null-hypothesis scenarios
  simulated. Every point estimate is reported with its Morris et al.
  Table 6 Monte Carlo standard error. Code and a full ADEMP audit
  trail accompany the simulation.
bibliography: references.bib
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
---
```{r include=F, echo=F}
rm(list = ls())
options(
  knitr.kable.NA = "", dplyr.summarise.inform = FALSE,
  dplyr.print_max = 1e9, knitr.table.format = "latex"
)
library(tidyverse)
library(lme4)
library(knitr)
library(kableExtra)
library(mvtnorm)

opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.width = 4.5, fig.height = 3.5,
  results = "asis", dev = "pdf",
  cache.path = "cache/"
)
# source(here::here("analysis/scripts/zz.tools.R"))
options(scipen = 1, digits = 3)
```

# Introduction

Multicenter randomized clinical trials (RCTs) represent the
standard design for confirmatory evaluation of therapeutic
interventions. By enrolling participants across multiple
clinical sites, these trials achieve practical recruitment
targets while broadening the population base for
generalization of findings [@iche9; @localio2001adjustments].
However, the multicenter structure introduces several analytic
complexities that, if mishandled, can distort inference about
treatment effects. We address four interrelated statistical
issues that arise in the design and analysis of multicenter
RCTs with continuous outcomes, and we present simulation
evidence bearing on each in turn.

## Modeling site effects

The first question concerns how site (or center) effects
should be incorporated into the primary analysis. Three
broad strategies exist: ignoring site differences entirely,
treating sites as fixed effects, and treating sites as
random effects drawn from a population of potential sites.
Each strategy carries distinct assumptions and implications
for the validity and efficiency of treatment effect
estimation [@senn1998controversies; @jones1998comparison].

@localio2001adjustments provided an influential overview of
the problem, identifying three analytic challenges inherent
to multicenter data: within-center correlation of outcomes,
confounding by center, and effect modification across
centers. The choice between fixed and random effects models
has been debated extensively. Fixed effects models condition
on the observed sites and make no distributional assumptions
about site effects, but they consume degrees of freedom
rapidly when the number of sites is large relative to total
sample size [@pickering2007analysis]. Random effects models
treat site effects as realizations from a normal distribution,
estimating only a variance component rather than individual
site parameters, and thereby preserve statistical efficiency
[@bates2015lme4]. The ICH E9 guideline recommends that mixed
models are 'especially relevant when the number of sites is
large' [@iche9].

@chu2011comparing conducted a comprehensive simulation
comparing six analytic approaches for continuous outcomes
across varying numbers of centers, center sizes, and
intraclass correlation coefficients (ICCs). All methods
yielded unbiased point estimates, but ignoring centers
inflated the standard error and reduced power when
within-center correlation was present. The mixed-effects
model attained nominal coverage and near-optimal power
across nearly all configurations. Generalized estimating
equations (GEE) underestimated the standard error when the
number of centers was small.

@kahanmorris2013continuous extended these findings,
demonstrating that random center effects performed at least
as well as fixed effects in all scenarios examined, and
substantially outperformed them when the number of patients
per center was small. @islam2024accounting confirmed these
patterns in a recent replication study with both continuous
and binary outcomes, reinforcing the recommendation for
random effects models as a default analytic strategy.

## Failure to adjust for stratification variables

The second issue concerns the consequences of ignoring
stratification variables, including site, in the analysis.
When randomization is stratified by center (as is standard
in multicenter trials), failing to adjust for the
stratification factor in the analysis leads to biased
inference. @kahanImproperAnalysisTrials2012a demonstrated
through simulation that an unadjusted analysis of a trial
randomized with stratified blocks produces standard errors
that are biased upward, confidence intervals that are too
wide, type I error rates below the nominal level, and a
loss of statistical power. This paradoxical conservatism
arises because the stratified randomization induces
negative correlation between treatment groups within strata;
ignoring this correlation inflates variance estimates.

@kahanAdjustingMultiplePrognostic2013 extended this work
to trials with multiple stratification factors, finding
that adjusting for the main effects of all stratification
variables was generally sufficient without requiring
adjustment for all cross-classified strata. These findings
align with the ICH E9 recommendation that 'the
statistical model used for analysis should reflect the
restriction imposed by the randomisation procedure'
[@iche9], and with current regulatory guidance on covariate
adjustment more broadly [@fda2023covariate].

## Treatment-by-site interaction

The third concern is heterogeneity of the treatment effect
across sites. Even under a common protocol, eligible
participants, site practices, and investigator experience
may vary, producing genuine treatment-by-site interaction
[@fedorov2005misleading]. @senn1998controversies argued
that some degree of treatment-by-center interaction is
inevitable in practice and that a rational approach to
planning multicenter trials should anticipate it.

The ICH E9 guideline recommends first fitting a model
without the interaction term to estimate the main treatment
effect, and then examining heterogeneity as a secondary
analysis. However, the power to detect treatment-by-site
interaction is typically low, so a non-significant
interaction test does not establish homogeneity
[@sennlewis2019treatment]. @jones1998comparison compared
fixed-effects weighted estimators, unweighted estimators,
and random effects estimators in the presence of varying
degrees of treatment-by-center interaction. The fixed
effects weighted estimator performed well across a range of
interaction magnitudes, while unweighted estimators could
be inefficient when site sizes differed markedly.

When treatment-by-site interaction is modeled as random
(i.e., the treatment slope varies across sites), the mixed
model estimates both the average treatment effect and the
variance of treatment effects across sites, providing a
natural summary of heterogeneity
[@neuhaus2006separating].

## Imbalanced site sizes

The fourth issue, closely related to the preceding three,
is whether imbalance in site sizes affects inference.
Multicenter trials frequently exhibit highly unequal
enrollment across sites, with a few large sites contributing
a disproportionate share of participants. @senn1998controversies
noted that a rational approach to planning multicenter
trials leads naturally to unequal site sizes, because sites
with higher recruitment capacity will enroll more patients
within the trial's fixed time window.

@ruvuna2004unequal introduced a coefficient of imbalance
to quantify the power loss attributable to unequal center
sizes and showed that substantial imbalance can reduce
power when an unweighted (Type III) analysis is used.
@vierron2009design formalized this through a design
effect framework, demonstrating that the efficiency loss
depends on both the ICC and a statistic quantifying the
heterogeneity of group distributions across centers. When
the ICC is positive, balanced allocation is optimal, but
moderate imbalance imposes only a modest efficiency
penalty. The key risk emerges when site size is
informative (that is, when larger sites systematically
differ from smaller sites in patient characteristics or
treatment delivery), potentially introducing bias into
the treatment effect estimate
[@fedorov2005misleading; @agresti2000strategies].

## Present study

The existing literature provides clear theoretical guidance
but limited integrated simulation evidence spanning all
four issues simultaneously. We present a Monte
Carlo simulation that jointly varies the analytic method
(ignore site, fixed site effects, random site effects),
the presence or absence of treatment-by-site interaction,
balanced versus imbalanced site sizes, and the magnitude
of site-level variance. The simulation evaluates bias,
empirical standard error, model-based standard error,
power, and coverage probability of 95\% confidence
intervals for the average treatment effect across all
factorial combinations of these design features.

# Methods

## ADEMP structure

Following Morris, White, and Crowther [-@morris2019using], the
simulation is reported under the ADEMP framework.

- **Aims.** Compare three analytic strategies for handling site
  heterogeneity in multicenter RCTs (ignore site, fixed site effects,
  random site effects) across scenarios varying the number of sites,
  site variance, treatment-by-site interaction, and site-size
  balance, and separately assess whether these strategies control
  the Type I error rate under a true null treatment effect.
- **Data-generating mechanism.** Detailed in the next subsection.
- **Estimand.** The overall treatment effect $\delta$ on the
  outcome scale, in the simulation-performance sense of Morris et al.
  (a data-generating parameter with a known true value), not in the
  ICH E9(R1) sense of a fully specified population/variable/
  intercurrent-event/summary-measure estimand [-@iche9r1].
- **Methods.** OLS ignoring site, OLS with fixed site effects, and a
  linear mixed model with random site effects via \texttt{lme4::lmer}.
  All three methods are tested and interval-estimated using a
  $t$-reference distribution: the OLS residual degrees of freedom for
  the ignore-site and fixed-site methods, and $N$ minus the number of
  fixed-effect parameters for the random-site method. The latter is a
  naive approximation, not a Kenward-Roger or Satterthwaite
  denominator-df correction (see Future research); it is used
  consistently for both the $p$-value and the 95\% CI half-width for
  that method, replacing an earlier implementation in which the
  $p$-value used this $t$-approximation but the CI used a fixed
  $z = 1.96$ half-width regardless of sample size or method.
- **Performance measures.** Bias, empirical SE, MSE, mean model SE,
  power at $\alpha = 0.05$ (equivalently, Type I error at
  $\delta = 0$ scenarios), coverage of 95\% CIs, convergence rate,
  and singular/boundary-fit rate, each reported with its Monte Carlo
  SE per Morris Table 6. Non-convergence (a hard error from
  \texttt{lmer}) and singular/boundary fits (a successful fit at the
  edge of the random-effects covariance parameter space, which
  \texttt{lme4} flags as a warning rather than an error) are tracked
  as two distinct first-class outcomes and reported per scenario and
  method, rather than treating a fit that raised any warning as
  simply non-convergent or, conversely, absorbing a boundary fit
  silently into "converged" with no record that it was a boundary
  solution.

## Data generating model

For each simulated trial, data are generated from a
two-level model with patients nested within sites.
Let $Y_{ij}$ denote the continuous outcome for patient $j$
at site $i$:

$$Y_{ij} = \alpha_i + (\delta + \gamma_i) T_{ij} + \epsilon_{ij}$$

where $\alpha_i \sim N(0, \sigma^2_{\alpha})$ is the
random site intercept, $\delta = 0.30$ is the true average
treatment effect, $\gamma_i \sim N(0, \sigma^2_{\gamma})$
is the random treatment-by-site interaction,
$T_{ij} \in \{0, 1\}$ is the treatment indicator
(balanced within each site via permuted blocks), and
$\epsilon_{ij} \sim N(0, \sigma^2_e = 1)$ is the
residual error.

## Parameter values

The simulation crosses the following factors:

- **Number of sites:** $K = 10$ (few) vs $K = 30$ (many)
- **Site-level variance:** $\sigma^2_{\alpha} \in \{0, 0.10, 0.50\}$
  (none, moderate, large)
- **Treatment-by-site interaction:**
  $\sigma^2_{\gamma} \in \{0, 0.04\}$ (none vs present)
- **Site size balance:** balanced ($n_i = 20$ for all $i$)
  vs imbalanced (site sizes drawn from a
  $\text{Gamma}(0.8, 0.8/20)$ distribution, yielding a
  coefficient of variation $\approx 1.1$)

The true treatment effect is $\delta = 0.30$ across all
scenarios, representing a moderate standardized effect size
of 0.30 standard deviations.

## Analytic methods

Each simulated dataset is analyzed by three methods:

1. **Ignore site:** $Y_{ij} = \beta_0 + \beta_1 T_{ij} + e_{ij}$
   (ordinary least squares, no site term)
2. **Fixed site effects:**
   $Y_{ij} = \beta_0 + \beta_1 T_{ij} + \sum_{k=2}^{K} \beta_k I(i=k) + e_{ij}$
   (site indicators as fixed effects)
3. **Random site effects:**
   $Y_{ij} = \beta_0 + \beta_1 T_{ij} + u_i + e_{ij}$
   where $u_i \sim N(0, \sigma^2_u)$ (linear mixed model
   via REML, fit with \texttt{lme4})

## Simulation procedure

```{r sim-run, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/scripts/sim_multicenter.R"))}
source(here::here("analysis/scripts/sim_multicenter.R"))

scenarios <- tibble::tribble(
  ~label, ~n_sites, ~n_per_site, ~balanced,
  ~sigma_site, ~sigma_trt_site, ~true_trt, ~sigma_e,
  "K10_noVar_bal",     10, 20, TRUE,  0.0, 0.0, 0.3, 1,
  "K10_modVar_bal",    10, 20, TRUE,  sqrt(0.10), 0.0, 0.3, 1,
  "K10_hiVar_bal",     10, 20, TRUE,  sqrt(0.50), 0.0, 0.3, 1,
  "K10_modVar_imbal",  10, 20, FALSE, sqrt(0.10), 0.0, 0.3, 1,
  "K10_hiVar_imbal",   10, 20, FALSE, sqrt(0.50), 0.0, 0.3, 1,
  "K10_modVar_int",    10, 20, TRUE,  sqrt(0.10), 0.2, 0.3, 1,
  "K10_hiVar_int",     10, 20, TRUE,  sqrt(0.50), 0.2, 0.3, 1,
  "K10_hiVar_int_imb", 10, 20, FALSE, sqrt(0.50), 0.2, 0.3, 1,
  "K30_noVar_bal",     30, 20, TRUE,  0.0, 0.0, 0.3, 1,
  "K30_modVar_bal",    30, 20, TRUE,  sqrt(0.10), 0.0, 0.3, 1,
  "K30_hiVar_bal",     30, 20, TRUE,  sqrt(0.50), 0.0, 0.3, 1,
  "K30_modVar_imbal",  30, 20, FALSE, sqrt(0.10), 0.0, 0.3, 1,
  "K30_hiVar_imbal",   30, 20, FALSE, sqrt(0.50), 0.0, 0.3, 1,
  "K30_modVar_int",    30, 20, TRUE,  sqrt(0.10), 0.2, 0.3, 1,
  "K30_hiVar_int",     30, 20, TRUE,  sqrt(0.50), 0.2, 0.3, 1,
  "K30_hiVar_int_imb", 30, 20, FALSE, sqrt(0.50), 0.2, 0.3, 1,
  # Null-hypothesis arm (true_trt = 0): a reduced factorial subset
  # of the eight alternative-hypothesis conditions above (dropping
  # the treatment-by-site-interaction rows, since sigma_trt_site
  # has no bearing on size when the average effect is null and
  # dropping them keeps the added runtime modest), needed to
  # estimate empirical Type I error rather than only coverage of a
  # CI centered on a nonzero effect (see Results, "Type I error
  # under the null").
  "H0_K10_noVar_bal",    10, 20, TRUE,  0.0, 0.0, 0.0, 1,
  "H0_K10_modVar_bal",   10, 20, TRUE,  sqrt(0.10), 0.0, 0.0, 1,
  "H0_K10_hiVar_bal",    10, 20, TRUE,  sqrt(0.50), 0.0, 0.0, 1,
  "H0_K10_hiVar_imbal",  10, 20, FALSE, sqrt(0.50), 0.0, 0.0, 1,
  "H0_K30_noVar_bal",    30, 20, TRUE,  0.0, 0.0, 0.0, 1,
  "H0_K30_modVar_bal",   30, 20, TRUE,  sqrt(0.10), 0.0, 0.0, 1,
  "H0_K30_hiVar_bal",    30, 20, TRUE,  sqrt(0.50), 0.0, 0.0, 1,
  "H0_K30_hiVar_imbal",  30, 20, FALSE, sqrt(0.50), 0.0, 0.0, 1
)

# Morris, White, and Crowther (2019) §4.1: pin RNGkind and set the
# seed ONCE at the start of the program. `run_simulation()` no longer
# accepts a `seed` argument; it consumes RNG state from the single
# stream established here.
RNGkind("L'Ecuyer-CMRG")
set.seed(20260310)

raw_results <- run_simulation(scenarios, n_sim = 1500)
sim_summary <- summarize_simulation(raw_results)

# The alternative-hypothesis scenarios (true_trt = 0.30) drive the
# main results table and figures, as in the original design; the
# null-hypothesis scenarios are reported separately (Results,
# "Type I error under the null") because a CI's coverage of a
# nonzero true effect is not the same quantity as the test's Type I
# error rate under the null, and merging the two arms into one table
# or one set of scenario-indexed figures would conflate them.
sim_summary_alt <- dplyr::filter(sim_summary, true_trt > 0)
sim_summary_null <- dplyr::filter(sim_summary, true_trt == 0)
raw_results_alt <- dplyr::filter(raw_results, true_trt > 0)
```

Each scenario is replicated $R = 1{,}500$ times. The Monte Carlo SE
of an estimated coverage or Type I error probability $p$ is
$\text{MCSE} = \sqrt{p(1-p)/R}$ (Morris, White, and Crowther 2019,
Table 6); evaluated at the nominal $p = 0.95$ (worst case among the
coverage targets considered here), $R \geq 1{,}320$ is required for
MCSE $\leq 0.6$ percentage points, and $R = 1{,}500$ gives
MCSE $\approx 0.56$ pp, satisfying the target with a margin.
Treatment is assigned via balanced allocation within each site (equal
numbers of treatment and control participants, or within $\pm 1$ when
$n_i$ is odd). All random number generation uses a fixed seed
(\texttt{set.seed(20260310)}) and the \texttt{L'Ecuyer-CMRG} RNG
for reproducibility. The `sim-run` chunk's cache is keyed on the MD5
hash of \texttt{analysis/scripts/sim\_multicenter.R} in addition to
the chunk's own text, so a change to the simulation functions
themselves (not just to this chunk) correctly invalidates the cached
result rather than silently reusing output from a prior version of
the simulation code.

# Results

## Headline findings

```{r headline-07, include=FALSE}
# Pull key numbers from the in-memory `sim_summary_alt` /
# `sim_summary_null` produced by the `sim-run` chunk above. If the
# sim has not run (e.g., cache invalid), these computations will
# fail at knit time.
hl_random <- sim_summary_alt |>
  dplyr::filter(method == "random_site") |>
  dplyr::summarise(
    power_min = min(power), power_max = max(power),
    cov_min = min(coverage), cov_max = max(coverage),
    bias_abs_max = max(abs(bias)),
    singular_max = max(singular_rate)
  )
hl_ignore <- sim_summary_alt |>
  dplyr::filter(method == "ignore_site") |>
  dplyr::summarise(
    cov_min = min(coverage), cov_max = max(coverage),
    power_min = min(power)
  )
hl_conv <- sim_summary_alt |>
  dplyr::summarise(conv_min = min(convergence_rate))
hl_t1 <- sim_summary_null |>
  dplyr::group_by(method) |>
  dplyr::summarise(
    t1_min = min(power), t1_max = max(power), .groups = "drop"
  )
hl_t1_random <- dplyr::filter(hl_t1, method == "random_site")
hl_t1_ignore <- dplyr::filter(hl_t1, method == "ignore_site")
```

We observe that across `r nrow(sim_summary_alt) / 3` alternative-
hypothesis scenarios ($\delta = 0.30$) and $R = 1{,}500$
replications per scenario, the random-site mixed model attained
power between
`r sprintf('%.2f', hl_random$power_min)` and
`r sprintf('%.2f', hl_random$power_max)`, with coverage of
the 95\% confidence interval ranging from
`r sprintf('%.3f', hl_random$cov_min)` to
`r sprintf('%.3f', hl_random$cov_max)` and absolute bias
never exceeding
`r sprintf('%.3f', hl_random$bias_abs_max)`
(Morris Table 6 Monte Carlo SEs are given in parentheses next to
the power and coverage point estimates in Table 1). The fixed-site
estimator closely
tracked the random-site estimator. Ignoring site entirely produced
overcoverage (up to
`r sprintf('%.3f', hl_ignore$cov_max)`) and correspondingly
lower power (as low as
`r sprintf('%.2f', hl_ignore$power_min)`) in scenarios with
non-trivial between-site variance. The lowest convergence rate
(hard \texttt{lmer} errors only) observed across any scenario or
method was `r sprintf('%.3f', hl_conv$conv_min)`; however, the
random-site model's *singular/boundary* fit rate reached as high as
`r sprintf('%.3f', hl_random$singular_max)`
in the low- and no-site-variance scenarios, where the true
random-intercept variance is at or near the parameter-space
boundary. A singular fit still returns a usable point estimate, but
its variance-component estimate is degenerate; see "Type I error
under the null" below and the Discussion for the implications. Under
the null-hypothesis scenarios ($\delta = 0$; Table 2), the empirical
Type I error rate for the random-site method ranged from
`r sprintf('%.3f', hl_t1_random$t1_min)` to
`r sprintf('%.3f', hl_t1_random$t1_max)`, and for the ignore-site
method from `r sprintf('%.3f', hl_t1_ignore$t1_min)` to
`r sprintf('%.3f', hl_t1_ignore$t1_max)`, against a nominal
$\alpha = 0.05$.

## Summary of performance metrics

```{r results-table}
sim_display <- sim_summary_alt |>
  mutate(
    scenario = str_replace_all(scenario, "_", " "),
    method = case_when(
      method == "ignore_site" ~ "Ignore",
      method == "fixed_site" ~ "Fixed",
      method == "random_site" ~ "Random"
    ),
    convergence = sprintf("%.1f\\%%", 100 * n_converged / n_total),
    singular = sprintf("%.1f\\%%", 100 * singular_rate),
    power_mcse = sprintf("%.3f (%.3f)", power, mcse_power),
    coverage_mcse = sprintf("%.3f (%.3f)", coverage, mcse_coverage)
  ) |>
  select(
    Scenario = scenario, Method = method,
    Bias = bias, `Emp. SE` = empirical_se,
    `Model SE` = mean_model_se,
    `Power (MCSE)` = power_mcse, `Coverage (MCSE)` = coverage_mcse,
    Conv. = convergence, Sing. = singular
  )

kable(
  sim_display,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  escape = FALSE,
  caption = "Simulation results under the alternative hypothesis
    ($\\delta = 0.30$): bias, standard error, power, and coverage
    by scenario and analytic method, each with its Morris et al.
    (2019) Table 6 Monte Carlo SE in parentheses where applicable.
    R = 1500 replications per scenario. `Conv.' is the rate of
    successful \\texttt{lmer} fits (errors only); `Sing.' is the
    rate of singular/boundary fits among converged random-site
    fits (not applicable to Ignore/Fixed). Type I error results
    for the corresponding null-hypothesis ($\\delta = 0$) scenarios
    are reported separately in Table 2.",
  linesep = ""
) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 7
  )
```

## Bias

```{r fig-bias, fig.cap="Bias of the treatment effect estimate across scenarios and analytic methods. Dashed line at zero indicates no bias.", fig.width=5, fig.height=3.5}
sim_summary_alt |>
  mutate(
    method = factor(method,
      levels = c("ignore_site", "fixed_site",
                 "random_site"),
      labels = c("Ignore", "Fixed", "Random")
    ),
    k = ifelse(grepl("K10", scenario), "K = 10", "K = 30"),
    scenario_short = str_remove(scenario, "K\\d+_")
  ) |>
  ggplot(aes(x = scenario_short, y = bias,
             shape = method, linetype = method,
             group = method)) +
  geom_point(size = 2) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey50") +
  facet_wrap(~k) +
  scale_shape_manual(values = c(16, 17, 15)) +
  scale_linetype_manual(values = c("solid", "dashed",
                                   "dotted")) +
  labs(x = NULL, y = "Bias",
       shape = "Method", linetype = "Method") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                   size = 6))
```

## Power and coverage

```{r fig-power, fig.cap="Power (left) and coverage probability (right) across scenarios. Dashed lines indicate the nominal 5\\% significance level (power panel) and 95\\% coverage target.", fig.width=6, fig.height=3.5}
p_data <- sim_summary_alt |>
  mutate(
    method = factor(method,
      levels = c("ignore_site", "fixed_site",
                 "random_site"),
      labels = c("Ignore", "Fixed", "Random")
    ),
    k = ifelse(grepl("K10", scenario), "K = 10", "K = 30"),
    scenario_short = str_remove(scenario, "K\\d+_")
  )

p1 <- p_data |>
  ggplot(aes(x = scenario_short, y = power,
             shape = method, linetype = method,
             group = method)) +
  geom_point(size = 2) +
  geom_line() +
  facet_wrap(~k) +
  scale_shape_manual(values = c(16, 17, 15)) +
  scale_linetype_manual(values = c("solid", "dashed",
                                   "dotted")) +
  labs(x = NULL, y = "Power",
       shape = "Method", linetype = "Method") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                   size = 5))

p2 <- p_data |>
  ggplot(aes(x = scenario_short, y = coverage,
             shape = method, linetype = method,
             group = method)) +
  geom_point(size = 2) +
  geom_line() +
  geom_hline(yintercept = 0.95, linetype = "dashed",
             color = "grey50") +
  facet_wrap(~k) +
  scale_shape_manual(values = c(16, 17, 15)) +
  scale_linetype_manual(values = c("solid", "dashed",
                                   "dotted")) +
  labs(x = NULL, y = "Coverage",
       shape = "Method", linetype = "Method") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                   size = 5))

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Impact of imbalanced site sizes

```{r fig-imbalance, fig.cap="Comparison of balanced versus imbalanced site sizes for the random effects method. Panels show empirical standard error (left) and power (right).", fig.width=5.5, fig.height=3}
bal_data <- raw_results_alt |>
  filter(method == "random_site", !is.na(est)) |>
  mutate(
    balanced = case_when(
      grepl("bal$", scenario) ~ "Balanced",
      grepl("imbal$|imb$", scenario) ~ "Imbalanced",
      TRUE ~ NA_character_
    ),
    base = str_remove(scenario, "_bal$|_imbal$|_imb$"),
    k = ifelse(grepl("K10", scenario), "K = 10", "K = 30")
  ) |>
  filter(!is.na(balanced))

bal_summary <- bal_data |>
  group_by(base, k, balanced) |>
  summarize(
    emp_se = sd(est),
    power = mean(pval < 0.05),
    .groups = "drop"
  )

p3 <- bal_summary |>
  ggplot(aes(x = base, y = emp_se,
             fill = balanced)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("grey85", "grey45")) +
  labs(x = NULL, y = "Empirical SE", fill = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                   size = 6))

p4 <- bal_summary |>
  ggplot(aes(x = base, y = power,
             fill = balanced)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("grey85", "grey45")) +
  labs(x = NULL, y = "Power", fill = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,
                                   size = 6))

gridExtra::grid.arrange(p3, p4, ncol = 2)
```

## Type I error under the null

Coverage of a 95\% CI centered on the true alternative
($\delta = 0.30$, Table 1) is not the same quantity as the Type I
error rate of the associated test under a true null effect
($\delta = 0$), which is the quantity most directly relevant to the
stratified-randomization literature motivating this study
(@kahanImproperAnalysisTrials2012a; @kahanAdjustingMultiplePrognostic2013).
Table 2 reports the empirical rejection rate at $\alpha = 0.05$ for
the reduced null-hypothesis scenario subset described in Methods.

```{r results-table-null}
sim_display_null <- sim_summary_null |>
  mutate(
    scenario = str_remove(scenario, "^H0_") |>
      str_replace_all("_", " "),
    method = case_when(
      method == "ignore_site" ~ "Ignore",
      method == "fixed_site" ~ "Fixed",
      method == "random_site" ~ "Random"
    ),
    t1_mcse = sprintf("%.3f (%.3f)", power, mcse_power),
    coverage_mcse = sprintf("%.3f (%.3f)", coverage, mcse_coverage),
    singular = sprintf("%.1f\\%%", 100 * singular_rate)
  ) |>
  select(
    Scenario = scenario, Method = method,
    Bias = bias, `Type I error (MCSE)` = t1_mcse,
    `Coverage (MCSE)` = coverage_mcse, Sing. = singular
  )

kable(
  sim_display_null,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  escape = FALSE,
  caption = "Empirical Type I error under the null hypothesis
    ($\\delta = 0$) at nominal $\\alpha = 0.05$, by scenario and
    analytic method, with Morris et al. (2019) Table 6 Monte Carlo
    SE in parentheses. R = 1500 replications per scenario. `Sing.'
    is the random-site singular/boundary-fit rate among converged
    fits.",
  linesep = ""
) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 8
  )
```

All three methods maintain Type I error close to the nominal 5\%
level across the null-hypothesis scenarios, including the
high-site-variance and imbalanced conditions, and including the
scenarios in which the random-site model's singular-fit rate is
highest. This is consistent with the alternative-hypothesis
coverage results in Table 1 and provides direct evidence -- rather
than an inference from coverage of a shifted CI -- that none of the
three methods inflates the false-positive rate under the conditions
simulated here. It does not, by itself, establish that the
$t$-based reference distribution used for the random-site method
(Methods, "Analytic methods") is correctly calibrated in general;
the boundary/singular fits flagged in Table 1 and Table 2 return a
point estimate and standard error that are numerically valid but
whose sampling distribution at the parameter-space boundary is not
guaranteed to match the naive $t$ approximation used here, and this
simulation's null-arm results should not be read as validating that
approximation outside the specific scenarios simulated.

# Discussion

The simulation results may be organized around the four
research questions motivating this study.

**Modeling site effects.** Consistent with
@chu2011comparing and @kahanmorris2013continuous, we find
that the random effects model attained nominal or near-
nominal coverage and the highest power across nearly all
scenarios. The fixed effects model performed comparably
when the number of sites was small ($K = 10$) but lost
efficiency with $K = 30$ due to the large number of
nuisance parameters. Ignoring site effects produced
unbiased point estimates but inflated standard errors when
site-level variance was present, leading to conservative
inference and reduced power. These findings support the
random effects model as a robust default for multicenter
RCTs, as recommended by @sennlewis2019treatment and
@iche9. A caveat attaches specifically to the no- and
low-site-variance scenarios: the random-intercept model's
variance component is frequently estimated at the
parameter-space boundary there (singular/boundary fit rates
up to `r sprintf('%.1f', 100 * hl_random$singular_max)`\%;
Table 1), which is expected behavior for REML at a
near-zero true variance component but means the reported
coverage and power for the random-site method in those
scenarios reflect a mix of interior and boundary solutions
rather than a uniformly well-behaved asymptotic regime.

**Failure to adjust for stratification.** The 'ignore
site' method effectively represents a failure to adjust
for the stratification variable (site) used in
randomization. The overcoverage and power loss observed
under this method confirm the findings of
@kahanImproperAnalysisTrials2012a: when randomization is
stratified by center, the analysis must account for
center to obtain valid inference. The magnitude of the
power penalty depends on the ICC: with no site-level
variance, ignoring site is harmless, but as
$\sigma^2_{\alpha}$ increases, the penalty becomes
substantial.

**Treatment-by-site interaction.** Introducing random
treatment-by-site interaction ($\sigma^2_{\gamma} = 0.04$)
increased the variability of treatment effect estimates
across replications for all methods. The random effects
model, which does not explicitly model this interaction
in the fitted specification, nonetheless maintained
reasonable coverage because the inflated variability was
partially absorbed into the residual. However, the
simulation did not fit a model with random slopes, which
would be the correct specification under this generating
mechanism. @jones1998comparison and
@fedorov2005misleading have argued that some degree of
treatment-by-center interaction is ubiquitous, and that
the power to detect it is typically inadequate. In our
view, researchers should plan for its presence rather
than test for its absence.

**Imbalanced site sizes.** The comparison of balanced
versus imbalanced site sizes reveals that for the random
effects model, moderate imbalance (CV $\approx 1.1$)
produces only a modest increase in empirical standard
error and a correspondingly small decrease in power. This
finding is consistent with @ruvuna2004unequal and
@vierron2009design, who showed that the efficiency loss
from unequal site sizes is generally small unless the
imbalance is extreme. The more concerning scenario is when
site size is confounded with site characteristics -- a
possibility not simulated here but discussed by
@fedorov2005misleading and @neuhaus2006separating.
Under informative site sizes, where larger sites
systematically differ from smaller sites, all methods
may produce biased estimates of the population-average
treatment effect.

**Limitations.** We should note that this simulation is
restricted to continuous outcomes analyzed with normal-
theory models.
Binary and time-to-event outcomes raise additional
complications, including separation in fixed effects
logistic regression with many small sites
[@kahanAccountingCentreeffectsMulticentre2014;
@agresti2000strategies; @pickering2007analysis]. The
simulation does not incorporate dropout, non-compliance,
or informative site sizes. The random effects model
fitted here includes only a random intercept; a random
slopes model would be more appropriate when
treatment-by-site interaction is expected. The
null-hypothesis (Type I error) arm covers a reduced subset
of eight of the sixteen alternative-hypothesis scenarios,
omitting the treatment-by-site-interaction conditions; Type
I error under a null average effect combined with genuine
site-level interaction variance is not assessed here and is
a natural extension once the random-slopes model above is
implemented. The random-site method's $p$-values and CIs
both use an $N$-minus-fixed-effects degrees of freedom that
is not a Kenward-Roger or Satterthwaite correction (Methods,
"Analytic methods"); the null-arm results in Table 2 show
this approximation was not anti-conservative in the
scenarios simulated, but that is not a general guarantee,
particularly for the boundary/singular fits discussed above.

# Future research

Several directions warrant further investigation:

1. **Random slopes models.** Extending the simulation to
   include a random treatment slope
   ($\delta + \gamma_i$ with $\gamma_i$ estimated) would
   clarify when and how much the random slopes model
   outperforms the random intercept model under genuine
   treatment heterogeneity.

2. **Non-normal outcomes.** Binary and count outcomes
   require generalized linear mixed models or GEE, and
   the relative performance of methods may differ from
   the continuous case, particularly with sparse data per
   site [@agresti2000strategies;
   @kahanAccountingCentreeffectsMulticentre2014].

3. **Informative site sizes.** If site size is associated
   with site-level treatment effect (e.g., expert centers
   enroll more patients and also deliver more effective
   treatment), standard methods may yield biased
   population-average estimates. Inverse-probability
   weighting by site or joint modeling of enrollment and
   outcome merit investigation
   [@neuhaus2006separating].

4. **Small-sample corrections.** The current implementation
   uses a naive $N$-minus-fixed-effects degrees of freedom
   for the random-site method's $p$-values and CIs (Methods,
   "Analytic methods"), which performed adequately in the
   null-arm scenarios simulated here (Table 2) but is not a
   general substitute for a proper small-sample correction.
   Kenward-Roger or Satterthwaite denominator degrees of
   freedom corrections (available in \texttt{lmerTest})
   should be substituted and evaluated across the same
   scenario grid, particularly for the $K = 10$, high-
   site-variance conditions where few sites most threaten
   the adequacy of any df approximation.

5. **Bayesian approaches.** Hierarchical Bayesian models
   offer partial pooling of site-specific treatment
   effects, which may outperform frequentist random
   effects models when the number of sites is very small
   [@jones1998comparison].

6. **Sample size planning.** Tools that incorporate the
   design effect from multicenter structure, including
   anticipated ICC and site size imbalance, into
   prospective sample size calculations would improve
   trial planning [@vierron2009design;
   @sverdlovSelectingRandomizationMethod2024].

# Data and code availability

All data reported here are synthetic, generated by
`analysis/scripts/sim_multicenter.R`; no external or human-participant
data are used. The simulation and reporting code, including this
document's source, are version-controlled in the repository containing
this report. The simulation is fully reproducible from the fixed seed
and RNG kind documented in Methods ("Simulation procedure"); the exact
scenario grid used to produce every table and figure in this report is
defined in the `sim-run` code chunk of `report.Rmd`.

# Conflict of interest

The author declares no conflict of interest. No external funding
supported this work.

# References

## Morris et al. (2019) ADEMP Compliance

We audited this simulation study against the reporting
standards proposed by Morris, White, and Crowther
[-@morris2019using]. The original audit is recorded at
`docs/morris-audit-2026-04-17.md`; a status update following the
2026-08 revision described below is appended to that file.

**Verdict:** Partially compliant (updated).

**Resolved since the original 2026-04-17 audit:**

- Monte Carlo SEs are now computed for every performance estimate in
  `summarize_simulation()` (`mcse_bias`, `mcse_empirical_se`,
  `mcse_mse`, `mcse_model_se`, `mcse_power`, `mcse_coverage`,
  `mcse_convergence`, `mcse_singular_rate`) and are displayed
  alongside the corresponding point estimates for power and coverage
  in Table 1 and Table 2.
- `R = 1500` is now justified by an explicit Monte Carlo SE
  derivation (Methods, "Simulation procedure").
- Non-convergence is no longer conflated with dropped-and-ignored
  replications; `n_converged` and `convergence_rate` are reported
  explicitly, and paired comparisons across methods are preserved
  because every method's row is retained (as `NA` on non-convergence)
  rather than filtered out.
- `RNGkind("L'Ecuyer-CMRG")` is pinned immediately before the single
  `set.seed()` call.
- Per-replicate `.Random.seed` values are captured by
  `run_simulation()` and attached as the `rng_states` attribute of
  its return value (not yet persisted to a sidecar file on disk,
  which remains an open item below).

**Gaps identified in the 2026-08 revision, not present in the
original audit:**

- Singular/boundary `lmer` fits were, prior to this revision, silently
  absorbed into "converged" by the warning-suppression logic in
  `fit_methods()`; this has been fixed by recording `singular` and
  `conv_warning` as explicit outcomes, but the fix is new as of this
  revision and has not yet been audited against Morris §5.1 on a
  wider scenario set than the one simulated here.
- Coverage previously used a fixed $z = 1.96$ half-width, inconsistent
  with the $t$-based $p$-values used for power; both now use the same
  per-method degrees of freedom (Methods, "Analytic methods"), but
  that shared degrees-of-freedom choice is itself a naive
  approximation for the random-site method, not a Kenward-Roger or
  Satterthwaite correction (Future research, item 4).
- The simulation previously provided no null-hypothesis arm and could
  not support a Type I error claim; a reduced null-arm scenario
  subset has been added (Results, "Type I error under the null"), but
  it does not cover the treatment-by-site-interaction conditions
  (Discussion, "Limitations").

**Still open:**

- Per-replicate RNG states are captured in memory but not persisted
  to `analysis/data/derived_data/rng_states.rds` as originally
  recommended, so a specific failing replication from a completed run
  cannot yet be re-run from disk without re-executing the whole
  simulation up to that point.
- `inst/tinytest/test_basic.R` now exercises `generate_trial_data()`
  and `summarize_simulation()` (see the package test suite), but does
  not yet cover `fit_methods()`'s singular-fit detection logic against
  a constructed non-singular case, which would strengthen confidence
  in the newly added `singular`/`conv_warning` columns.


```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/(Library/CloudStorage/)?Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
